---
title: "Tutorial 2" 
subtitle: "Visualizing Financial Time Series and Checking Normality"
author: "Barry Quinn"
date: "Last updated: `r Sys.Date()`"
format:
    html:
        code-fold: true
        embed-resources: true
        toc: true
        toc-depth: 3
        code-tools:
             source: https://github.com/advanced-financial-data-analytics/tutorial2
number-sections: true
editor: visual
execute:
    echo: true
    warning: false  
    message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(tidyquant)
library(fpp2)
library(tsfe)     # Contains class datasets if desired
library(ggpmisc)  # For annotation and peak/valley
library(DT)       # Optional for interactive tables
```

# Introduction

This tutorial helps you develop data exploration skills for financial time series. You will learn to: 

1. Use time-series objects in R to organize your data. 
2. Create effective visual arguments: time plots, seasonal plots, and subseries plots. 
3. Understand distribution properties: Compare your data to the normal distribution (histogram overlays, QQ-plots). 
4. Use autocorrelation (ACF) to detect patterns like trend or seasonality. 
5. Draw practical conclusions about normality assumptions in finance.

## Financial Context

- Financial time series (e.g., stock returns) are often assumed to be normally distributed for simplicity in risk/portfolio models. This assumption underlies key theories like the Capital Asset Pricing Model (CAPM) and Black-Scholes option pricing.
- Empirical evidence, however, shows fat-tailed distributions (more extreme events than normal), volatility clustering (periods of high/low volatility tend to persist), and other departures from normality.
- These deviations from normality can lead to underestimation of market risk, especially during crisis periods.
- Visualization is a first step in diagnosing these issues and understanding the true nature of financial risk.

# Data Setup

We'll illustrate with a monthly log returns series for Glencore ("GLEN.L"). Feel free to swap in other tickers or use your class data from tsfe.

```{r}
# Using tidyquant to get monthly log returns from 2018
glen <- tq_get("GLEN.L", from = "2018-01-01") %>%
tq_transmute(select = adjusted,
mutate_fun = monthlyReturn,
type = "log",
col_rename = "log_return")

# Let's convert to a ts object (freq=12 for monthly)
glen_ts <- ts(glen$log_return, frequency = 12, start = c(2018,1))

# Peek at the data
head(glen, 10)
```

> You'll see two columns: date and log_return. The glen_ts object is for using forecast-style time-series functions.

# Time-Series Visualisation

## Simple Time Plot

```{r}
autoplot(glen_ts) +
labs(
title = "Glencore Monthly Log Returns",
x = "Time",
y = "Log Return"
) +
theme_minimal()
```

> **Economic Interpretation:**
> - Look for clustering of volatility (periods where returns are consistently more volatile)
> - Major spikes often correspond to significant market events or company news
> - The overall pattern can indicate regime changes in the company's risk profile

## Seasonal Plot

While monthly equity returns often don't have a strong seasonal pattern, let's demonstrate the function:

```{r}
ggseasonplot(glen_ts, year.labels = TRUE, year.labels.left = TRUE) +
labs(
title = "Seasonal Plot: Glencore Monthly Returns",
y = "Log Return",
x = "Month"
)
```

> **Statistical & Economic Context:**
> - While pure seasonal patterns are rare in equity returns (would violate market efficiency),
> - Some sectors show seasonal effects (e.g., retail stocks around holidays)
> - Trading volume often has stronger seasonality than returns
> - If you see patterns, consider if they're economically exploitable after transaction costs

## Subseries Plot

```{r}
ggsubseriesplot(glen_ts) +
labs(
title = "Subseries Plot: Glencore Monthly Returns",
y = "Log Return"
)
```

> Subseries grouping by month helps highlight any recurring monthly patterns.

# Normality Checks

## Histogram with Overlaid Normal

```{r}
glen %>%
    ggplot(aes(x = log_return)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill="skyblue", alpha = 0.8) +
    stat_function(
        fun = dnorm,
        args = list(mean = mean(glen$log_return),
        sd   = sd(glen$log_return)),
        color = "red",
        size = 1
        ) +
    labs(
    title = "Glencore Log Returns vs. The Normal Distribution",
    x = "Log Return",
    y = "Density"
    ) +
    theme_minimal()
```

> **Statistical Insights:**
> - Fat tails indicate higher probability of extreme events than predicted by normal distribution
> - Asymmetry (skewness) might suggest different probabilities of positive vs negative returns
> - The shape often reflects the market regime (crisis vs stable periods show different distributions)

## QQ-Plot

```{r}
glen %>% ggplot(aes(sample = log_return)) + 
    stat_qq(distribution = qnorm) + 
    stat_qq_line(distribution = qnorm, color="red") + 
    labs( title = "QQ Plot of Glencore Log Returns", 
           x = "Theoretical Quantiles (Normal)", 
           y = "Sample Quantiles" ) + 
    theme_minimal()
```

**Interpretation:** 

- Points above the line in the right tail indicate more extreme positive returns than expected under normality
- Points below the line in the left tail indicate more extreme negative returns
- The pattern often reveals asymmetric risk (e.g., crashes being more severe than booms)
- This has important implications for risk management and option pricing

## Shapiro-Wilk Test

```{r}
shapiro.test(glen$log_return)
```

- Null hypothesis: The data are normally distributed. 
- If p-value \< 0.05, we usually reject normality. 
- Caution: Large sample sizes can make tiny deviations appear significant.

# Autocorrelation & Seasonal Patterns

## Autocorrelation Function (ACF)

```{r}
ggAcf(glen_ts) + labs(title = "ACF: Glencore Monthly Log Returns")
```

**ACF Interpretation:** 
- Significant autocorrelation challenges the Efficient Market Hypothesis
- Positive autocorrelation suggests momentum effects (trending)
- Negative autocorrelation suggests mean reversion
- The decay pattern of ACF can inform trading strategy horizons
- Remember that statistical significance doesn't always equal economic significance (after costs)

## Lag Plot

```{r}
gglagplot(glen_ts)
```

Each small scatterplot checks log_return~t~ vs. log_return~(t - k)~

# Exercises

Change the Ticker: 

Choose a different stock (e.g. "AMZN", "BARC.L" for Barclays, etc.) and replicate steps 3–5. 

2. Discuss Normality: 
- Does the histogram or QQ-plot indicate fat tails? 
- Why might normality be a poor assumption for daily or monthly returns? 

3. Autocorrelation: 
- Are you seeing any significant correlation at lag=1 or lag=2? 
- If so, what might that imply about "efficient market" assumptions? 

4. Seasonality:
- If you do see a monthly pattern, can you guess economic reasons behind it?

# Conclusion

:::{.callout-tip}
### Key Takeaways: 
- Visual arguments are vital: time plots, QQ-plots, and ACF often reveal critical patterns that summary statistics obscure
- Normality assumptions in finance are convenient but dangerous - risk models need to account for fat tails
- Market efficiency doesn't require perfect randomness - small predictabilities may exist but be hard to profit from
- The interaction between statistical patterns and economic reasoning is crucial for financial analysis
- These tools form the foundation for more advanced models in risk management and trading
:::
